---
title: "R Notebook"
---




```{r}
library(SingleCellExperiment)
library(tidyverse)
```


```{r}
set.seed(1)
```


Data was downloaded from https://singlecell.broadinstitute.org/single_cell/study/SCP1663

Unfortunately, this cannot be automated, so you will manually need to download the files

* predictive_variable_slideseq_j20_rep1.rds
* predictive_variable_slideseq_j20_rep2.rds
* predictive_variable_slideseq_j20_rep3.rds
* predictive_variable_slideseq_j20_rep4.rds
* SpatialRNA_cropped_slideseq_j20_rep1.zip
* SpatialRNA_cropped_slideseq_j20_rep2.zip
* SpatialRNA_cropped_slideseq_j20_rep3.zip
* SpatialRNA_cropped_slideseq_j20_rep4.zip

Unzip the zip files. Those will contain RDS files (without the file extension). Lastly, move the files into the `data/cable_spatial_plaque_data` folder.

```{r}
rep1 <- readRDS("../data/cable_spatial_plaque_data/SpatialRNA_cropped_slideseq_j20_rep1")
rep2 <- readRDS("../data/cable_spatial_plaque_data/SpatialRNA_cropped_slideseq_j20_rep2")
rep3 <- readRDS("../data/cable_spatial_plaque_data/SpatialRNA_cropped_slideseq_j20_rep3")
rep4 <- readRDS("../data/cable_spatial_plaque_data/SpatialRNA_cropped_slideseq_j20_rep4")

plaque_dens_rep1 <- readRDS("../data/cable_spatial_plaque_data/predictive_variable_slideseq_j20_rep1.rds")
plaque_dens_rep2 <- readRDS("../data/cable_spatial_plaque_data/predictive_variable_slideseq_j20_rep2.rds")
plaque_dens_rep3 <- readRDS("../data/cable_spatial_plaque_data/predictive_variable_slideseq_j20_rep3.rds")
plaque_dens_rep4 <- readRDS("../data/cable_spatial_plaque_data/predictive_variable_slideseq_j20_rep4.rds")
```




```{r, paged.print = FALSE}
common_gene_names <- intersect(intersect(rownames(rep1@counts), rownames(rep2@counts)), intersect(rownames(rep3@counts), rownames(rep4@counts)))

merged_mat <- cbind(
  rep1@counts[common_gene_names, ], 
  rep2@counts[common_gene_names, ], 
  rep3@counts[common_gene_names, ], 
  rep4@counts[common_gene_names, ]
)
colnames(merged_mat) <- c(paste0(colnames(rep1@counts), "-rep1"),
                          paste0(colnames(rep2@counts), "-rep2"),
                          paste0(colnames(rep3@counts), "-rep3"),
                          paste0(colnames(rep4@counts), "-rep4"))

sce <- SingleCellExperiment(list(counts = merged_mat), colData = bind_rows(rep1@coords, rep2@coords, rep3@coords, rep4@coords) %>% magrittr::set_rownames(NULL))
sce$sample <- rep(paste0("mouse_", 1:4), times = c(ncol(rep1@counts), ncol(rep2@counts), ncol(rep3@counts), ncol(rep4@counts)))
logcounts(sce) <- transformGamPoi::shifted_log_transform(sce)
sce$plaque_density <- c(plaque_dens_rep1, plaque_dens_rep2, plaque_dens_rep3, plaque_dens_rep4)

# Remove lowly expressed genes
sce <- sce[rowSums(counts(sce)) > 50, ]
sce
```


Add gene ids
```{r, paged.print = FALSE}
eg <- clusterProfiler::bitr(rownames(sce), fromType = "SYMBOL", toType = "ENSEMBL", OrgDb="org.Mm.eg.db") %>%
  summarize(ENSEMBL = dplyr::first(ENSEMBL), .by = SYMBOL)
  
rowData(sce) <- data.frame(symbol = rownames(sce)) %>%
  tidylog::left_join(dplyr::rename(eg, symbol = SYMBOL, gene_id = ENSEMBL)) 
```




Add cell type labels to each cell

```{r}
brain_atlas_data_file <- "../data/cable_spatial_plaque_data/metacells.BrainCellAtlas_Saunders_version_2018.04.01.RDS"
brain_atlas_metadata_file <- "../data/cable_spatial_plaque_data//annotation.BrainCellAtlas_Saunders_version_2018.04.01.RDS"
if(! file.exists(brain_atlas_data_file) || ! file.exists(brain_atlas_metadata_file)){
  # See http://dropviz.org/ for more information
  data_url <- "https://storage.googleapis.com/dropviz-downloads/static/metacells.BrainCellAtlas_Saunders_version_2018.04.01.RDS"
  annotation_url <- "https://storage.googleapis.com/dropviz-downloads/static/annotation.BrainCellAtlas_Saunders_version_2018.04.01.RDS"
  
  download.file(data_url, brain_atlas_data_file)
  download.file(annotation_url, brain_atlas_metadata_file)
}
metacells_saunders <- readRDS(brain_atlas_data_file)
metacells_annotation <- readRDS(brain_atlas_metadata_file)
brain_ref <- SingleCellExperiment(list(counts = metacells_saunders), colData = metacells_annotation)
logcounts(brain_ref) <- transformGamPoi::shifted_log_transform(counts(brain_ref))
hippocampus_ref <- brain_ref[, brain_ref$tissue == "HC"]

length(intersect(rownames(brain_ref), rownames(sce)))
length(rownames(brain_ref))
length(setdiff(rownames(sce), rownames(brain_ref)))
# str(setdiff(rownames(sce), rownames(brain_ref)))

pred <- SingleR::SingleR(sce, ref = hippocampus_ref, labels = hippocampus_ref$class)
sce$cell_type <- pred$labels
sce$cell_type_lumped <- fct_lump(sce$cell_type, n = 5)

SingleR::plotScoreHeatmap(pred)
```



```{r}
hist(sce$plaque_density)
```


```{r, paged.print=FALSE}
qc_df <- scuttle::perCellQCMetrics(sce, subsets = list(Mito = str_starts(rowData(sce)$symbol, "mt-")))
hist(qc_df$subsets_Mito_percent, breaks = 100)
hist(qc_df$sum, breaks = 100)

sce_subset <- sce[, qc_df$sum > 200 & qc_df$sum <= 5000 & qc_df$subsets_Mito_percent < 20]
sce_subset
```



Calculate UMAP
```{r, paged.print=FALSE}
set.seed(1)
umap <- scater::calculateUMAP(sce_subset)
reducedDim(sce_subset, withDimnames = FALSE) <- umap
```




```{r}
cont2ordered <- function(x, n, from = min(x), to = max(x), add_label = TRUE, label_fmt = "%.2f"){
  groups <- seq(from, to, length.out = n+1)
  res <- rep(NA_integer_, length(x))
  labels <- rep(NA_character_, n)
  for(idx in seq_len(n)){
    if(idx == n){
      labels[idx] <- paste0("[", sprintf(label_fmt, groups[idx]), ",", sprintf(label_fmt, groups[idx+1]), "]")
      res[x >= groups[idx] & x <= groups[idx+1]] <- idx
    }else{
      labels[idx] <- paste0("[", sprintf(label_fmt, groups[idx]), ",", sprintf(label_fmt, groups[idx+1]), ")")
      res[x >= groups[idx] & x < groups[idx+1]] <- idx
    }
  }
  if(add_label){
    ordered(res, levels = seq_len(n), labels = labels)
  }else{
    res
  }
}

cont2ordered(seq(0, 30), n = 6, label_fmt = "%.f")
levels(cont2ordered(sce_subset$plaque_density, n = 10, label_fmt = "%.2f"))
```



```{r}
sce_subset$plaque_cluster <- cont2ordered(sce_subset$plaque_density, n = 10, label_fmt = "%.2f")
```


```{r}
as_tibble(colData(sce_subset)) %>%
  mutate(umap = umap) %>%
  ggplot(aes(x = umap[,1], y = umap[,2])) +
    geom_point(aes(color = sample), size = 0.3, stroke = 0) +
    guides(color = guide_legend(override.aes = list(size = 1)))

as_tibble(colData(sce_subset)) %>%
  mutate(umap = umap) %>%
  ggplot(aes(x = umap[,1], y = umap[,2])) +
    geom_point(aes(color = cell_type_lumped), size = 0.3, stroke = 0) +
    guides(color = guide_legend(override.aes = list(size = 1)))

as_tibble(colData(sce_subset)) %>%
  mutate(umap = umap) %>%
  ggplot(aes(x = umap[,1], y = umap[,2])) +
    geom_point(aes(color = plaque_density), size = 0.3, stroke = 0) +
    scale_color_viridis_b()
```


```{r}
set.seed(1)
fit <- lemur::lemur(sce_subset, design = ~ plaque_cluster, n_embedding = 30, test_fraction = 0.6)
set.seed(1)
fit <- lemur::align_harmony(fit)
```


```{r}
fit <- lemur::test_de(fit, cond(plaque_cluster = "[0.90,1.00]") - cond(plaque_cluster = "[0.00,0.10)"))
set.seed(1)
nei <- lemur::find_de_neighborhoods(fit, group_by = vars(plaque_cluster, sample), test_method = "edgeR")
```


```{r}
qs::qsave(sce_subset, "../data/cable_spatial_plaque_data/annotated_sce.qs")
```

```{r}
fit_small <- fit["Jun",]
nei_small <- nei %>% filter(name == "Jun")

qs::qsave(fit_small, "../output/cable_spatial_plaque_data/fit_small.qs")
qs::qsave(nei_small, "../output/cable_spatial_plaque_data/nei_small.qs")
```


# Session Info

```{r}
sessionInfo()
```


